\chapternonum{Introduction}
\label{chap:intro}


\vspace{5mm}

\section{What is \aevol{}?}
\label{sec:whatis}
\aevol{} is a digital genetics model: populations of digital organisms are
subjected to a process of selection and variation, which creates a
Darwinian dynamics.

By modifying the characteristics of selection (e.g. population size,
type of environment, environmental variations) or variation (e.g.
mutation rates, chromosomal rearrangement rates, types of
rearrangements, horizontal transfer), one can study experimentally the
impact of these parameters on the structure of the evolved organisms.
In particular, since \aevol{} integrates a precise and realistic model of
the genome, it allows for the study of structural variations of the
genome (e.g. number of genes, synteny, proportion of coding sequences).

The simulation platform comes along with a set of tools to help analyse
phylogenies and to measure many characteristics of the organisms and
populations along evolution.


\section{License}
\label{sec:license}
This program is free software: you can redistribute it and/or modify
it under the terms of the GNU General Public License as published by
the Free Software Foundation, either version 3 of the License, or
(at your option) any later version.

This program is distributed in the hope that it will be useful,
but WITHOUT ANY WARRANTY; without even the implied warranty of
MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
GNU General Public License for more details.

You should have received a copy of the GNU General Public License
along with this program.  If not, see <http://www.gnu.org/licenses/>.


\section{The \aevol{} Community}
\label{sec:comm}
\aevol{}'s primary ressource is its website \verb?http://www.aevol.fr/? where you shall find information about the project and its contributors.

To subscribe to the (low trafic) users' mailing lists, please visit \\\verb?http://lists.gforge.liris.cnrs.fr/mailman/listinfo/aevol-users?.

You may also want to report bugs and ask for new features to be implemented. \\To do so, simply write to \\\verb?aevol-bugs@lists.gforge.liris.cnrs.fr? or \\\verb?aevol-feat-request@lists.gforge.liris.cnrs.fr?


\clearemptydoublepage






